---
title: "DS8006 - Lab 3"
author: "Najlis, Bernardo"
date: "February 12, 2017"
output: pdf_document
---


## Step 1

Complete the '10,000 tweets' tutorial using twitteR searchTwitter function to load 1,000 tweets that mention @RyersonU.


```{r, message=F, warning=F}
library(ggplot2)
library(lubridate)
library(scales)
library(twitteR)
library(dplyr)

# Oauth authentication with API key and token
consumer_key <- "pY9vu9ZlKAHIVjO79cNAMH8zy"
consumer_secret <- "0TZZl6UW7UYDrt3sh27wB5qKZArrvhSHmw1aDsIz7xOa9vXCKv"
access_token <- "824392227749298177-UfnBjHAkOaJG6wRIvJye4gN6MOAWp8c"
access_secret <- "Byom7w6p6IUMcrwwKY0jhUCvHGBxMcC5fUZuOash0EHtX"

# Twitter oauth
setup_twitter_oauth(consumer_key, consumer_secret, access_token, access_secret)

# Get tweets using searchTwitter
ryeTweets  <-twListToDF( searchTwitter("@RyersonU", n=1000) )

# Add boolean feature isReply
ryeTweets$isReply <- ifelse((!is.na(ryeTweets$replyToSID) |
                              !is.na(ryeTweets$replyToSN) |
                              !is.na(ryeTweets$replyToUID)), TRUE, FALSE)
```
## Step 2

Revise the scripts from Step 1 to plot the following charts:
## Task 1

Number of Reply posts vs. other posts

```{r}

ggplot(ryeTweets, aes(ryeTweets$isReply)) +
  geom_bar(fill = "midnightblue") + 
  theme(legend.position="none", axis.title.x = element_blank()) +
  ylab("Number of tweets") + 
  ggtitle("Tweets by Type (Reply vs. Other)") +
  scale_x_discrete(labels=c("Other", "Reply tweets"))
```

Same plot modified to show percentage values
```{r}
ggplot(ryeTweets, aes(x = ryeTweets$isReply)) +
  geom_bar(aes(y=(..count..) / sum(..count..)), fill = "midnightblue") + 
  theme(legend.position="none", axis.title.x = element_blank()) +
  scale_y_continuous(labels=percent) +
  labs(title = "Tweets by Type (Reply vs. Other)", y="Percentage of tweets") + 
  scale_x_discrete(labels=c("Other", "Reply tweets"))
```

## Task 2

Number of Retweets vs other posts
```{r}

ggplot(ryeTweets, aes(ryeTweets$isRetweet)) +
  geom_bar(fill = "midnightblue") + 
  theme(legend.position="none", axis.title.x = element_blank()) +
  ylab("Number of tweets") + 
  ggtitle("Tweets by Type (Retweet vs. Other)") +
  scale_x_discrete(labels=c("Other", "Retweets"))
```

Same plot modified to show percentage values.
```{r}
ggplot(ryeTweets, aes(x = ryeTweets$isRetweet)) +
  geom_bar(aes(y=(..count..) / sum(..count..)),fill = "midnightblue") + 
  theme(legend.position="none", axis.title.x = element_blank()) +
  scale_y_continuous(labels=percent) +
  labs(y="Number of tweets", title="Tweets by Type (Retweet vs. Other)") +
  scale_x_discrete(labels=c("Other", "Retweets"))
```

## Task 3

Plot comparing the distribution of the number of times a post was retweeted, for the original tweets vs retweets.

```{r}
ggplot(data = ryeTweets, aes(x = retweetCount)) +
  geom_histogram(aes(fill = ..count..), bins=20, binwidth = 1) +
  theme(legend.position = "none") +
  xlab("Retweet count") + ylab("Number of tweets") + ggtitle("Number of Retweets for original tweets vs. Retweets") +
  xlim(0,30) +
  scale_fill_gradient(low = "midnightblue", high = "aquamarine4") +
  facet_grid(isReply ~ .)
```

## Task 4

Propose a new char to visualize any other metadata field(s) available in your dataset.

```{r}
# Obtain top 10 users by number of tweets
top10 <- ryeTweets %>% count(screenName) %>% top_n(n=10)

ggplot(data= top10, aes(reorder(screenName, n),n)) +
 geom_bar(stat="identity", fill="midnightblue") +
  coord_flip() +
  labs(y="Number of Tweets", x="Screen Name",title="Most active Twitter Users")
```