---
title: "Lab 6 - Visualization of Dynamic Networks in R"
author: "Najlis, Bernardo"
date: "March 11, 2017"
output: html_document
---

## Data Extraction

1. Export data from Netlytic.org from previous lab.

## Visualization of Dynamic Networks in R

First approach, to create and display a basic network visualization.

```{r}
library(igraph)
edges <- read.table("net_JointAddress2.csv",header=T,sep=",")

g <- graph.edgelist(as.matrix(edges[,c(2,3)]),directed=T)
layout.old <- layout.graphopt(g)
 
plot(g,layout=layout.old,
        vertex.frame.color=V(g)$color,
        edge.width=1.5,
        asp=9/16)
```

Second attempt. This adjust the node size to reflect the number of connections (degree centrality) and also remove labels from nodes with less than 10 connections (chosen arbitrarily). We will also add the caption to the graph and will specify the size of edge arrows.

```{r}
plot(g,layout=layout.old,
        vertex.frame.color=V(g)$color,
        edge.width=1.5,
        asp=9/16,
        vertex.size= 1 + 1.5*log(graph.strength(g)),
        vertex.label=ifelse(degree(g)>10,V(g)$name,NA),
        vertex.label.color= "black",
        vertex.label.font=1,
        vertex.label.cex=2,
edge.arrow.size=0.1,
main="Dynamic Network Visualization"
)
```


```{r}
png(file="img/test.png", width=1600,height=900,bg = "#F1F1F5")
 
plot(g,layout=layout.old,
        vertex.frame.color=V(g)$color,
        edge.width=1.5,
        asp=9/16,
        vertex.size= 1 + 1.5*log(graph.strength(g)),
        vertex.label=ifelse(degree(g)>10,V(g)$name,NA),
        vertex.label.color= "black",
        vertex.label.font=1,
        vertex.label.cex=2,
edge.arrow.size=0.5,
main="Dynamic Network Visualization"
)
dev.off()
```

The following script (adapted from Moro’s code) is an extension of the previous script. Each edge in the network file has a timestamp value showing when the edge was created. The script uses this information to generate multiple snapshots of the network; each snapshot representing the network at a different point in time.

```{r}

E(g)$time <- edges[,4]
#remove self-loops
g <- simplify(g, remove.multiple = FALSE, remove.loops = TRUE)
step <- 3
E(g)$weight <- ifelse(E(g)$time < step,1,0)
layout.old <- layout.graphopt(g,niter=100,spring.length=E(g)$weight)
png(file="img/net%03d.png", width=1600,height=900,bg = "#F1F1F5")
total_time <- max(E(g)$time)
delta <- 0.5
nsteps <- max(E(g)$time)
for(step  in seq(3,total_time,delta)){  
  E(g)$weight <- ifelse(E(g)$time < step,1,0)
  E(g)$color <- ifelse(E(g)$time < step,"gray",rgb(0,0,0,0))
  V(g)$color <- ifelse(graph.strength(g)==0,rgb(0,0,0,0),"#3476A8")
  layout.new <- layout.graphopt(g,niter=10,start=layout.old,spring.length=E(g)$weight,max.sa.movement=1)
  plot(g,layout=layout.new,
        vertex.frame.color=V(g)$color,
        edge.width=1.5,
        asp=9/16,
        vertex.size= 1 + 1.5*log(graph.strength(g)),
        vertex.label=ifelse(degree(g)>10,V(g)$name,NA),
        vertex.label.color= "black",
        vertex.label.font=1,
        vertex.label.cex=2,
edge.arrow.size=0.5,
main="Dynamic Network Visualization"
)
  layout.old <- layout.new
}
dev.off()
```




